---
title: "Short analysis report"
author: "Alex Litovchenko"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Short analysis report}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 6,
  fig.height = 4
)
```

This vignette shows a minimal analysis report: fit a model, print fixed effects, plot fitted vs observed, and show residual ACF and Q-Q.

## 1. Fit model(s)

```{r fit}
library(tidyILD)
d <- ild_simulate(n_id = 8, n_obs_per = 10, irregular = TRUE, seed = 101)
x <- ild_prepare(d, id = "id", time = "time", gap_threshold = 7200)
x <- ild_center(x, y)
fit <- ild_lme(y ~ y_bp + y_wp + (1 | id), data = x, ar1 = FALSE, warn_no_ar1 = FALSE)
```

## 2. Tidy fixed-effects table

```{r tidy}
tidy_ild_model(fit)
```

With cluster-robust standard errors (requires the **clubSandwich** package):

```{r tidy_robust, eval = requireNamespace("clubSandwich", quietly = TRUE)}
tidy_ild_model(fit, se = "robust", robust_type = "CR2")
```

## 3. Fitted vs observed

```{r fitted_plot, fig.alt = "Fitted vs observed"}
ild_plot(fit, type = "fitted")
```

## 4. Residual diagnostics: ACF and Q-Q

```{r diag}
diag <- ild_diagnostics(fit, type = c("residual_acf", "qq"))
diag
```

```{r diag_plots, fig.alt = "Residual ACF and Q-Q", fig.show = "hold"}
plots <- plot_ild_diagnostics(diag)
plots$residual_acf
plots$qq
```

## With AR1 (nlme)

Same workflow with residual autocorrelation (nlme path). If the AR1 model fails to converge on this small example, we show the code only; in practice use more data or a different seed.

```{r fit_ar1}
fit_ar1 <- tryCatch(
  ild_lme(y ~ y_bp + y_wp, data = x, random = ~ 1 | id, ar1 = TRUE),
  error = function(e) NULL
)
if (!is.null(fit_ar1)) {
  tidy_ild_model(fit_ar1)
} else {
  message("AR1 fit did not converge on this run; use ild_lme(..., ar1 = TRUE) with your data.")
}
```

```{r diag_ar1, eval = !is.null(fit_ar1)}
if (!is.null(fit_ar1)) {
  diag_ar1 <- ild_diagnostics(fit_ar1, type = c("residual_acf", "qq"))
  diag_ar1
}
```

## Time-varying effects (TVEM)

When the effect of a predictor may change over time (e.g. stronger in the morning), use `ild_tvem()` to fit a GAM with a smooth in time and a time-varying coefficient. The plot shows how the effect of the predictor varies over the study timeline; interpret the curve as the estimated effect at each time point (with uncertainty band).

```{r tvem}
set.seed(101)
d2 <- ild_simulate(n_id = 6, n_obs_per = 12, seed = 101)
d2$x <- rnorm(nrow(d2))
x2 <- ild_prepare(d2, id = "id", time = "time")
tv <- ild_tvem(x2, "y", "x", k = 5, re_id = TRUE)
ild_tvem_plot(tv)
```
