---
title: "From raw data to model with tidyILD"
author: "tidyILD authors"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{From raw data to model with tidyILD}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 6,
  fig.height = 4
)
```

This vignette walks through the full tidyILD pipeline: prepare data, inspect structure, apply within-between decomposition and lags, fit a mixed-effects model, and run diagnostics and plots.

## Simulate and prepare

```{r prepare}
library(tidyILD)
# Simulate simple ILD
d <- ild_simulate(n_id = 10, n_obs_per = 12, irregular = TRUE, seed = 42)
# Prepare: encode time structure and add .ild_* columns
x <- ild_prepare(d, id = "id", time = "time", gap_threshold = 7200)
```

## Inspect

```{r summary}
ild_summary(x)
ild_spacing_class(x)
```

## Within-person centering and lags

```{r center_lag}
x <- ild_center(x, y)
x <- ild_lag(x, y, mode = "gap_aware", max_gap = 7200)
```

## Fit a model

Without residual autocorrelation (lmer):

```{r lme_no_ar1}
fit0 <- ild_lme(y ~ 1 + (1 | id), data = x, ar1 = FALSE, warn_no_ar1 = FALSE)
```

With AR1 residual correlation (nlme):

```{r lme_ar1}
fit1 <- ild_lme(y ~ 1, data = x, ar1 = TRUE, correlation_class = "CAR1")
```

## Diagnostics and plots

```{r diagnostics}
diag <- ild_diagnostics(fit1, data = x)
names(diag$plot)
```

```{r plot_trajectory, fig.alt = "Trajectory plot"}
ild_plot(x, type = "trajectory", var = "y", max_ids = 5)
```

```{r plot_fitted, fig.alt = "Fitted vs observed"}
ild_plot(fit1, type = "fitted")
```

## Reproducibility

Use a fixed seed when simulating or fitting models so results can be recreated. The pipeline is deterministic for a given seed and data. When saving results (e.g. after [ild_lme()] or [ild_diagnostics()]), you can attach a reproducibility manifest and save a single bundle with [ild_manifest()] and [ild_bundle()]:

```{r reproducibility}
# Optional: build a manifest with scenario and seed, then bundle the fit for saving
manifest <- ild_manifest(seed = 42, scenario = ild_summary(x), include_session = FALSE)
bundle <- ild_bundle(fit1, manifest = manifest, label = "model_ar1")
# saveRDS(bundle, "run.rds")  # one file with result + manifest + label
names(bundle)
```
